---
title: "MCS_analysis"
author: "Faye Sanders"
date: "2026-01-08"
output: html_document
---

```{r setup, include=FALSE}
library(foreign)
library(dplyr)
library(mice)
library(reshape2)
library(psych)
library(ggplot2)
library(broom)
library(knitr)
library(kableExtra)
library(GGally)


## Load datasets
### Green Space Deciles at Wave 5 (age 11 - maps most closely to our age 12 in ALSPAC)
green_space <- read.spss(".../UKDA-8154-spss/spss/spss28/green_space_ward_deciles_mcs5.sav", to.data.frame = TRUE, use.value.labels = FALSE)

### Urban-rural classification at Wave 5 
urbanicity <- read.spss(".../UKDA-7464-spss/spss/spss28/mcs5_geographically_linked_data.sav", to.data.frame = TRUE, use.value.labels = FALSE)
urbanicity <- urbanicity[, c("EMORPCDE", "MCSID")] # ONS 2005 urban morphology code

## Recode so that higher urbanicity = higher score
table(urbanicity$EMORPCDE)
urbanicity$urban <- ifelse(urbanicity$EMORPCDE == 1, 3,
                ifelse(urbanicity$EMORPCDE == 3, 1,
                       urbanicity$EMORPCDE)) 
table(urbanicity$urban)

urbanicity$EMORPCDE <- NULL

### Cognition at Age 17 sweep (Wave 7)
cognition <- read.spss(".../UKDA-8682-spss/spss/spss28/mcs7_cm_cognitive_assessment.sav", to.data.frame = TRUE, use.value.labels = FALSE)

### Smoking at Age 17 sweep (Wave 7)
smoking <- read.spss(".../UKDA-8682-spss/spss/spss28/mcs7_cm_interview.sav", to.data.frame = TRUE, use.value.labels = FALSE)
smoking <- smoking[, c("GCSMOK00", "MCSID")]
smoking <- smoking %>% rename(smoking = GCSMOK00)

### Sex at Age 17 sweep (Wave 7)
sex <- read.spss(".../UKDA-8682-spss/spss/spss28/mcs7_hhgrid.sav", to.data.frame = TRUE, use.value.labels = FALSE)
sex <- sex[, c("GHCSEX00", "MCSID")]

#### only keep sex of the cohort member
sex_non_missing <- sex %>%
  filter(!is.na(GHCSEX00))

sex_missing_one <- sex %>%
  group_by(MCSID) %>%
  filter(all(is.na(GHCSEX00))) %>%  # IDs with no sex anywhere
  slice(1) %>%                      # keep just one row
  ungroup()


sex_clean <- bind_rows(sex_non_missing, sex_missing_one)
sex_clean <- sex_clean %>% rename(sex = GHCSEX00)

### Age at Age 17 sweep (Wave 7)
age <- read.spss(".../UKDA-8682-spss/spss/spss28/mcs7_cm_derived.sav", to.data.frame = TRUE, use.value.labels = FALSE)
age <- age[, c("GCMCS7AG", "MCSID")]
age <- age %>% rename(age = GCMCS7AG)

### SES measures
employment_homeless <- read.spss(".../UKDA-7464-spss/spss/spss28/mcs5_parent_interview.sav", to.data.frame = TRUE, use.value.labels = FALSE)
income <- read.spss(".../UKDA-7464-spss/spss/spss28/mcs5_family_derived.sav", to.data.frame = TRUE, use.value.labels = FALSE)

## SES ##
income <- income[, c("EOEDE000", "MCSID")]
employment_homeless <- employment_homeless[, c("EPHOMS00", "EPACQU0A",
                                               "EPACQU0B",
                                               "EPACQU0C",
                                               "EPACQU0D",
                                               "EPACQU0E",
                                               "EPACQU0F",
                                               "EPACQU0G",
                                               "EPACQU0H",
                                               "EPACQU0I",
                                               "EPACQU0J",
                                               "EPACQU0K",
                                               "EPACQU0L", "MCSID")]

## Auxilliary variables ## (Harmonized to best of ability with ALSPAC)
# child ethnicity
ethnicity <- read.spss(".../UKDA-4683-spss/spss/spss28/mcs1_cm_derived.sav", to.data.frame = TRUE, use.value.labels = FALSE)

ethnicity <- ethnicity[, c("ADCEEA00_R30", "MCSID")]

## recode to match coding of ethnicity in ALSPAC, 1 = white, 2 = non-white
ethnicity$ethnicity_child <- ifelse(is.na(ethnicity$ADCEEA00_R30), NA,
                     ifelse(ethnicity$ADCEEA00_R30 %in% c(1, 2), 1, 2))

ethnicity$ADCEEA00_R30 <- NULL

## check it worked
table(ethnicity$ethnicity_child)

# active travel 

active_travel <- read.spss(".../UKDA-8156-spss/spss/spss28/mcs6_parent_cm_interview.sav", to.data.frame = TRUE, use.value.labels = FALSE)

active_travel <- active_travel[, c("FPTRSC00", "FPTRHO00", "FPTRDI00", "FPTRTI00", "MCSID")]

active_travel$FPTRDI00 <- ifelse(active_travel$FPTRHO00 == 1, 
                       active_travel$FPTRSC00,   # set to FPTRSC00 if FPTRHO00 == 1
                       active_travel$FPTRDI00)   # otherwise keep original value

active_travel <- active_travel %>% rename(travel_to_school = FPTRSC00)
active_travel <- active_travel %>% rename(travel_from_school = FPTRDI00)
active_travel <- active_travel %>% rename(school_travel_time = FPTRTI00)

# Mums depression
mum_depression <- read.spss(".../UKDA-8156-spss/spss/spss28/mcs6_parent_interview.sav", to.data.frame = TRUE, use.value.labels = FALSE)

mum_depression <- mum_depression[, c("FPPHHO00", "FPPHRF00", "FPPHEE00", "FPPHWO00", "FPPHNE00", "MCSID")]

# Time at address
time_address <- read.spss(".../UKDA-8156-spss/spss/spss28/mcs6_parent_interview.sav", to.data.frame = TRUE, use.value.labels = FALSE)

time_address <- time_address[, c("FPMOAD00", "MCSID")]

percent_missing <- sum(is.na(time_address$FPMOAD00)) / length(time_address$FPMOAD00) * 100
 
percent_missing 

### Harmonize highest level of education with how it is defined in ALSPAC

employment_homeless <- employment_homeless %>%
  mutate(
    highest_education = case_when(
      EPACQU0A == 1 | EPACQU0B == 1 ~ 1,  # degree
      EPACQU0F == 1 ~ 2,                  # a-level
      EPACQU0G == 1 ~ 3,                  # o-level
      EPACQU0C == 1 | EPACQU0D == 1 | EPACQU0E == 1 | EPACQU0I == 1 ~ 4,  # vocational
      EPACQU0H == 1 ~ 5,                  # cse
      EPACQU0J == 1 | EPACQU0K == 1 | EPACQU0L == 1 ~ NA_real_
      # everything else = NA
    )
  )

employment_homeless <- employment_homeless[, c("highest_education", "EPHOMS00", "MCSID")]


# OECD Weekly income = equivalised (Adjusted income that accounts for household size and composition (OECD scale))

income <- income %>% rename(OECD_weekly_income = EOEDE000)
employment_homeless <- employment_homeless %>% rename(ever_homeless = EPHOMS00)

SES_merged <- merge(income, employment_homeless, by = "MCSID")

### Check coding (higher score = worse SES)
table(SES_merged$highest_education)
table(SES_merged$ever_homeless)

# Homeless needs reverse coding so 1=yes, 0=no
SES_merged <- SES_merged %>%
  mutate(
    ever_homeless_r = ifelse(ever_homeless == 2, 0,
                    ifelse(ever_homeless == 1, 1, NA))  # optional: NA for anything else
  )

table(SES_merged$ever_homeless_r)

green_space <- green_space[, c("EGREENSP", "MCSID")]

green_space <- green_space %>% rename(green_space = EGREENSP)

# 1 = gave correct answer, all else = incorrect answer
vars <- c(
  "GCNUM00","GCCHIC100","GCNAEX00","GCNAEG1","GCNAEG2",
  "GCNAEG3","GCNAAS0A","GCNAAS0B","GCNAAS0C","GCNAAS0D",
  "GCNAAS0E","GCNAAS0F","GCNAAS0G","GCNAAS0H","GCNAAS0I","GCNAAS0J")

## Create score counting no. of 1's (correct answers)
cognition <- cognition %>%
  mutate(
    cognitive_performance = rowSums(across(all_of(vars)) == 1, na.rm = TRUE)
  )

## Only keep what we need
cognition <- cognition[, c("cognitive_performance", "MCSID")]

# Remove duplicates in all our dataframes
SES_merged <- SES_merged[!duplicated(SES_merged$MCSID), ]
age        <- age[!duplicated(age$MCSID), ]
sex_clean  <- sex_clean[!duplicated(sex_clean$MCSID), ]
smoking$MCSID <- trimws(smoking$MCSID)
smoking    <- smoking[!duplicated(smoking$MCSID), ]
ethnicity    <- ethnicity[!duplicated(ethnicity$MCSID), ]
active_travel    <- active_travel [!duplicated(active_travel $MCSID), ]
mum_depression    <- mum_depression[!duplicated(mum_depression$MCSID), ]

# Remove trailing white space in IDs
mum_depression$MCSID <- trimws(mum_depression$MCSID)
active_travel $MCSID <- trimws(active_travel $MCSID)
ethnicity $MCSID <- trimws(ethnicity $MCSID)

# Merge cognition and green space data by ID
merged <- merge(cognition, green_space, by = "MCSID", all.y = TRUE)
merged <- merge(merged, urbanicity, by = "MCSID", all.y = TRUE)

# Now merge our covariates with this
merged_all <- merge(merged, SES_merged, by = "MCSID", all.x = TRUE)
merged_all <- merge(merged_all, age, by = "MCSID", all.x = TRUE)
merged_all <- merge(merged_all, sex_clean, by = "MCSID", all.x = TRUE)
merged_all <- merge(merged_all, smoking, by = "MCSID", all.x = TRUE)
merged_all <- merge(merged_all, ethnicity, by = "MCSID", all.x = TRUE)
merged_all <- merge(merged_all, active_travel, by = "MCSID", all.x = TRUE)
merged_all <- merge(merged_all, mum_depression, by = "MCSID", all.x = TRUE)

# Complete cases for green space
merged_all <- merged_all[!is.na(merged_all$green_space), ]


# Compare missingness levels across variables
missing_pct <- sapply(merged_all, function(x) {
  mean(is.na(x)) * 100
})

missing_pct

merged_all$highest_education <- NULL
merged_all$ever_homeless <- NULL
merged_all$ever_homeless_r <- NULL

# Remove what we no longer need
rm(cognition, employment_homeless, green_space, income, merged, SES_merged, sex, sex_clean, sex_missing_one, sex_non_missing, age, smoking)
gc()


## Load imputed data
mod_imp <- readRDS("....rds")

imputed1 <- complete(mod_imp, 1)

numeric_vars <- imputed1 %>% select(cognitive_performance, green_space, OECD_weekly_income, age, urban)

cor_matrix <- cor(numeric_vars, use = "pairwise.complete.obs")

cor_long <- melt(cor_matrix)
colnames(cor_long) <- c("Variable1", "Variable2", "Correlation")

# Main model

fit <- with(
  mod_imp,
  lm(cognitive_performance ~ green_space * urban + sex + OECD_weekly_income + age + smoking)) 

pooled <- pool(fit)

pooled_summary <- summary(pooled)

# meta-analysis

# main

fit_main <- with(
  mod_imp,
  lm(cognitive_performance ~ urban * green_space + sex + OECD_weekly_income + age + smoking)) 

pooled_main <- pool(fit_main)

pooled_summary_main <- summary(pooled_main)

# Sensitvity (no smoking)
fit2 <- with(
  mod_imp,
  lm(cognitive_performance ~ urban * green_space + sex + OECD_weekly_income + age)) 

pooled2 <- pool(fit2)

pooled_summary2 <- summary(pooled2)

# unadjusted
fit3 <- with(
  mod_imp,
  lm(cognitive_performance ~ urban)) 

pooled3 <- pool(fit3)

pooled_summary3 <- summary(pooled3)

```

## Variable Information

### Smoking at Age 17 Sweep (Wave 7)
####  1 = I have never smoked cigarettes, 2 = I have only ever tried smoking cigarettes once, 3 = I used to smoke sometimes but I never smoke a cigarette now, 4 = I sometimes smoke cigarettes now but I don't smoke as many as once a week, 5 = I usually smoke between one and six cigarettes a week, 6 = I usually smoke more than six cigarettes a week

### Green Space at Age 11 Sweep (Wave 5)
#### This is land-use percentile green space data. (https://doc.ukdataservice.ac.uk/doc/8154/mrdoc/pdf/mcs_green_space_deciles_user_guide_v3.pdf). We only want to keep LSOA Green Space as it's the finest-grain measure of green space (e.g., we don't want ward as its larger than an LSOA)

### Urbanicity at Age 14 (Wave 6)
### using the 2004 Urban/Rural definitions by the Office for National Statistics (Office for National Statistics, 2004). On the super output area (SOA) level, areas are classed as either ‘Urban (population > 10,000)’, ‘Town and fringe’ or ‘Village, hamlet or isolated dwellings’.

### Cogitive Performance at Age 17 Sweep (Wave 7)
#### the Number Analogies activity (GL Assessments). The activity assesses young people’s basic arithmetic knowledge and reasoning with numbers. Ten items were used from the Cognitive Abilities Test 3, Level H, Number Analogies test (David F Lohman, Robert L Thorndike, Elizabeth P Hagen, Adapted by Pauline Smith, Cres Fernandes and Steve Strand) which assessed the young person’s arithmetic knowledge and reasoning with numbers. The assessment was used and reduced from the original 20 items with permission of the owners GL Assessment. © David F Lohman, Robert L Thorndike, Elizabeth P Hagen, 2001. Reproduced by permission of GL Assessment. The assessment was timed, and cohort members had 6 minutes to work through the 10 questions in the question booklet. They were provided with pencil and paper if needed to work out answers.  Interviewers were told not to administer the assessment if the cohort member had a learning disability or serious behavioural problem (e.g. severe ADHD, autism), or was unable to respond in the required manner for the assessment (e.g. reading). Before beginning the assessment, interviewers were asked to confirm in CAI whether they received verbal consent from the cohort member to carry out the cognitive assessment

## Descriptives before imputation

```{r descriptive before, echo = FALSE}

describe(numeric_vars)
cat("Table 1: Distribution of Sex\n")
table(merged_all$sex)
cat("\nTable 2: Distribution of Smoking\n")
table(merged_all$smoking) 
cat("\nTable 2: Distribution of Ethnicity\n")
table(merged_all$ethnicity_child)

# Cognitive performance
ggplot(merged_all, aes(x = cognitive_performance)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Cognitive Performance (Before Imputation)", x = "Cognitive Performance", y = "Count")

# Green space
ggplot(merged_all, aes(x = green_space)) +
  geom_histogram(binwidth = 1, fill = "lightgreen", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Green Space (Before Imputation)", x = "Green Space Decile", y = "Count")

# Urbanicity
ggplot(merged_all, aes(x = urban)) +
  geom_histogram(binwidth = 1, fill = "plum", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Urbanicity (Before Imputation)", x = "Urbanicity (3 = >10,000, 2 = town, 1 = village/hamlet", y = "Count")

# OECD weekly income
ggplot(merged_all, aes(x = OECD_weekly_income)) +
  geom_histogram(fill = "orange", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Income (Before Imputation)", x = "OECD Weekly Income", y = "Count")

# Age
ggplot(merged_all, aes(x = age)) +
  geom_histogram(binwidth = 1, fill = "purple", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Age (Before Imputation)", x = "Age", y = "Count")

# Missingness across all variables
cat("\nTable 3: Percentage Missingness across all variables\n")
missing_pct

```

## Descriptives after imputation

```{r descriptive imputed, echo = FALSE}

describe(imputed1 %>% select(cognitive_performance, green_space, smoking, urban, OECD_weekly_income, age))
table(imputed1$sex)
table(imputed1$urban)
table(imputed1$smoking)
table(imputed1$ethnicity_child)

# Extract first imputed dataset
imputed1 <- complete(mod_imp, 1)

# Cognitive performance
ggplot(imputed1, aes(x = cognitive_performance)) +
  geom_histogram(binwidth = 1, fill = "skyblue", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Cognitive Performance (After Imputation)", x = "Cognitive Performance", y = "Count")

# Green space
ggplot(imputed1, aes(x = green_space)) +
  geom_histogram(binwidth = 1, fill = "lightgreen", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Green Space (After Imputation)", x = "Green Space Decile", y = "Count")

# Urbanicity
ggplot(imputed1, aes(x = urban)) +
  geom_histogram(binwidth = 1, fill = "plum", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Urbanicity (After Imputation)", x = "Urbanicity (3 = >10,000, 2 = town, 1 = village/hamlet", y = "Count")

# OECD weekly income
ggplot(imputed1, aes(x = OECD_weekly_income)) +
  geom_histogram(fill = "orange", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Income (After Imputation)", x = "OECD Weekly Income", y = "Count")

# Age
ggplot(imputed1, aes(x = age)) +
  geom_histogram(binwidth = 1, fill = "purple", color = "black") +
  theme_minimal() +
  labs(title = "Distribution of Age (After Imputation)", x = "Age", y = "Count")


```

## Correlations

```{r correlations, echo=FALSE}

ggplot(cor_long, aes(x = Variable1, y = Variable2, fill = Correlation)) +
  geom_tile(color = "white") +   # each square
  geom_text(aes(label = round(Correlation, 2)), color = "black", size = 4) + # add correlation numbers
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0, limit = c(-1,1)) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
  labs(title = "Correlation Matrix (After Imputation)", x = "", y = "")

vars_for_corr <- c(
  "age",
  "cognitive_performance",
  "green_space",
  "urban",
  "ethnicity_child",
  "smoking",
  "OECD_weekly_income",
  "sex"
)

ggpairs(
   imputed1[, vars_for_corr],

  # LOWER: scatterplots + smooth ellipse analogue
  lower = list(
    continuous = wrap(
      "points",
      alpha = 0.4,
      size = 1
    )
  ),

  # DIAGONAL: histogram + density
  diag = list(
    continuous = wrap(
      "densityDiag",
      fill = "#D55E00",
      alpha = 0.6
    )
  ),

  # UPPER: Pearson correlations
  upper = list(
    continuous = wrap(
      "cor",
      method = "pearson",
      size = 4
    )
  )
) +
  theme_minimal() +
  labs(title = "Correlations After Imputation")

```

# Analyses

```{r analysis, echo=FALSE}

pooled_summary

```

# Sensitivity analyses (no smoking)

```{r sensitivity, echo=FALSE}

pooled_summary2

```

# results for meta-analysis

```{r main, echo=FALSE}

pooled_summary_main

pooled_summary2

pooled_summary3 

```


# Interaction Plot

```{r interaction, include=FALSE}

library(ggplot2)
library(dplyr)

imputed1 <- complete(mod_imp, 1)

# Make urban a labelled factor
imputed1 <- imputed1 %>%
  mutate(
    urban_f = factor(
      urban,
      levels = c(1, 2, 3),
      labels = c("Village/hamlet", "Town/fringe", "Urban >10,000")
    )
  )


```

```{r interaction plot, echo=FALSE}

ggplot(imputed1, aes(x = green_space,
                     y = cognitive_performance,
                     colour = urban_f)) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(
    x = "Green space (decile)",
    y = "Cognitive performance",
    colour = "Urbanicity",
    title = "Interaction between Green Space and Urbanicity"
  ) +
  theme_minimal()

```
